Ensemble Stability of Biological Networks

### 2014-09-25 paper.tex
update section on dynamical systems

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph, $G$, to the interdependencies, and a vector field $V \in \Gamma T(\prod_i M_i)$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
having connectivity $\connectivity$ equal to the number of edges of the graph, in this case $\connectivity = 3$. For a general system the directed graph $G_X$ that describes the manner in which each of the variables depends upon one another is given by the adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}.
   \right.
\end{displaymath} For the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

In order to discuss how different parts of the system influence each
other, we shall generalize our definition to include open systems.  We
do this by having some of our variables be control variables whose
values are given as specified functions of time rather than determined
by the dynamics.  Given several such open systems, we may combine them
to produce a larger system by setting the control variables of a
subsystem equal to the dynamical variables of another system.  For
details and a discussion of the operad structure of composition of
open systems see \cite{Vagner2014}.  Taking this notion to its logical
extreme, one can dissect a system into a collection of one open system
for each dynamical variable.  However, for most purposes, such a
decomposition would be uninteresting and what we instead want is a
decomposition into relatively self-contained modules.  In the section
on connected components, we shall propose a way of accomplishing this
based upon the topology of the system graph.

In order to accommodate the possibility of uncertainty in our
modelling, we will generalize our notion of a system on a network to
that of a system with random parameters.  Again, this will involve
three steps. First, we provide a measure space $S, \mathcal{B}(S)$
which represents the values over which our parameters can vary. Next,
instead of a single vector field, we consider a family of vector
fields parameterized by this space. That is to say, we introduce a
mapping $V \colon S \to \Gamma T(\prod_i M_m)$.  (Here $\Gamma$
denotes the space of sections of the tangent bundle.)  As before, for
each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu \colon
\mathcal{B}(S) \to [0,1]$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

@@

A quality which is of interest to us is robustness or structural
stability \cite{Smale1967}, which requires the determination of
whether or not a given dynamical system that is determined to be
stable remains stable under a perturbation to one or more of its
defining parameters.  We will quantify this as the probability that,
if some property holds for a set of parameters, it will continue to
hold if we make a random perturbation about those parameter values.
To do so, we will introduce, in addition to the probability
distribution $\mu$ described above, a family of probability
distributions $\mu' \colon S \times \mathcal{B}(S) \to [0,1]$ such
that $\mu'(x,y)$ encodes the conditional probability that a system
with a parameter value $x$ will have its parameter values changed to
$y$ under a perturbation.  For our biological models, we will be
interested in large perturbations rather than the small or
infinitessimal perturbations usually considered in the theory of
dynamical systems.  Then, given a binary random variable $q$,
we define its robustness as the following conditional proability:
\begin{align}\label{eq:robustness}
  R (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\end{align}

@@

A graph can be decomposed into a natural collection of \emph{modules} called strongly connected components. A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc} along with their adjacency matrices and hierarchy diagrams (to be defined below).

>>>>

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph, $G$, to the interdependencies, and a vector field $V$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
having connectivity $\connectivity$ equal to the number of edges of the graph, in this case $\connectivity = 3$. For a general system the directed graph $G_X$ that describes the manner in which each of the variables depends upon one another is given by the adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}.
   \right.
\end{displaymath} For the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

In order to accommodate the possibility of uncertainty in our
modeling, we will generalize our notion of a system on a network to
that of a system with random parameters.  Again, this will involve
three steps. First, we provide a measure space $S$
which represents the values over which our parameters can vary. Next,
instead of a single vector field, we consider a family of vector
fields parameterized by this space.  As before, for
each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure, $\mu$, which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

@@

A quality which is of interest to us is robustness or structural stability \cite{Smale1967}, which requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters.  We will quantify this as the probability that, if some property holds for a set of parameters, it will continue to hold if we make a random perturbation about those parameter values. To do so, we will introduce, in addition to the probability distribution $\mu$ described above, a family of probability distributions $\mu'$ such that $\mu'(x,y)$ encodes the conditional probability that a system with a parameter value $x$ will have its parameter values changed to $y$ under a perturbation.  For our biological models, we will be interested in large perturbations rather than the small or infinitesimal perturbations usually considered in the theory of dynamical systems.  Then, given a binary random variable $q$, we define its robustness as the following conditional probability:
\begin{align}\label{eq:robustness}
  R (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\end{align}

@@

Open systems are distinguished by having some of the variables be control variables whose values are given as specified functions of time rather than determined by the dynamics.  Given several such open systems, we may combine them to produce a larger system by setting the control variables of a subsystem equal to the dynamical variables of another system.  The composition and decomposition of open systems has algebraic structure equivalent to certain a kind of operad \cite{Vagner2014}.  Taking this notion to its logical extreme, one can dissect a system into a collection of one open system for each dynamical variable.  However, for most purposes, such a decomposition would be uninteresting and what we instead want is a decomposition into relatively self-contained modules.  One method of accomplishing this based upon the topology of the system graph is the decomposition into strongly connected components.

A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc} along with their adjacency matrices and hierarchy diagrams (to be defined below).
